% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrreprt}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{241,243,245}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother





\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}



 


\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{bookmark}{}{\usepackage{bookmark}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Quarto-SSF},
  pdfauthor={Jean Palate},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}


\title{Quarto-SSF}
\author{Jean Palate}
\date{2026-02-01}
\begin{document}
\maketitle

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}

\bookmarksetup{startatroot}

\chapter*{Index}\label{index}
\addcontentsline{toc}{chapter}{Index}

\markboth{Index}{Index}

\bookmarksetup{startatroot}

\chapter{}\label{section}

\bookmarksetup{startatroot}

\chapter*{References}\label{references}
\addcontentsline{toc}{chapter}{References}

\markboth{References}{References}

\phantomsection\label{refs}

\part{Introduction}

\chapter{Overview}\label{overview}

\section{Description of the general
model}\label{description-of-the-general-model}

The \texttt{general\ linear\ gaussian} state-space model can be written
in many different ways. The form considered in JD+ (\(\ge 3.0\)) is
presented below:

\[ y_t = Z_t \alpha_t + \epsilon_t,\quad \epsilon_t \sim N\left(0, \sigma^2 H_t\right),\quad t> 0\]

\[ \alpha_{t+1} = T_t \alpha_t + \mu_t, \quad \mu_t \sim N \left(0, \sigma^2 V_t \right),\quad t \ge 0 \]

\(y_t\) is the observation at period t, \(\alpha_t\) is the state
vector. \(\epsilon_t, \mu_t\) are assumed to be serially independent at
all leads and lags and independent from each other.\\
In the case of multi-variate models, \(y_t\) is a vector of
observations. However, in most cases, we will use the univariate
approach by considering the observations one by one (univariate handling
of multi-variate models).

The innovations of the state equation will be modeled as

\[ \mu_t = S_t \xi_t, \quad \xi_t \sim N\left( 0, \sigma^2 I\right) \]

In other words, \(V_t=S_t S_t'\)

The initial (\(\equiv t=0\)) conditions of the filter are defined as
follows:

\[ \alpha_{0} = b + B\delta + \mu_{0}\]

\(\delta\) is unknown or
\(\delta \sim N\left(0, \kappa \sigma^2 I \right)\), where \(\kappa\) is
arbitrary large. \(\mu_{0} \sim N\left(0, \sigma^2 P_*\right)\). So,
\(P_*\) is the variance of the stationary part of the initial state
vector and \(B\), which must be a full rank matrix, models the diffuse
part. We write \(BB'=P_\infty\). It should be noted that applying an
orthogonal transformation to \(B\) or multiplying it by a constant will
give exactly the same initialization. Moreover, when the initial state
is completely diffuse, \(\mu_0\) is not relevant and it can be safely
considered as \(0\). Finally, it should be noted that imposing that the
covariance of \(\delta\) is proportional to \(I\) is not an actual
restriction: in the hypothesis of a more general non singular matrix, we
just should pre-multiply \(\delta\) by the inverse of the Cholesky
factor of that matrix and post-multiply \(B\) by that factor to get an
equivalent initialization.

In the sequel, we will call ``stationary models'', the models with a
simple initialization (\$a\_0 = u\_0\$) and ``diffuse models'' (even
when \(\delta\) are considered as fixed unknown parameters), the models
with the general starting conditions.

The definition used in JD+ is quasi-identical to that of Durbin and
Koopman{[}1{]}.

In summary, apart from the scaling factor \(\sigma^2\) which can be
given or estimated, the model is completely defined by the following
quantities (possible default values are indicated in brackets):

\[ \mathbf{Z_t}, \mathbf{H_t} [=0] \]

\[ \mathbf{T_t}, \mathbf{V_t} [=S_t S_t'], \mathbf{S_t} [=Cholesky(V)] \]

\[ \mathbf{b}[=0], \mathbf{P_*} [=0], \mathbf{B} [=0], \mathbf{P_\infty} [=BB'] \]

It should be noted that all the indexes of arrays or matrices start at
\(0\) (Java and C++ convention). The following notations are also used
in all the pages related to the ssf framework:

\[ a_{t+k|t}=E\left(\alpha_{t+k} | y_0 \cdots y_{t}\right)\]

\[ P_{t+k|t}=cov\left(\alpha_{t+k} | y_0 \cdots y_{t}\right)\]

\[ a_{t+1}=a_{t+1|t}, \: P_{t+1}=P_{t+1|t} \] \[ 
\begin{align*}
y_t - E(y_t| y_0 \cdots y_t-1) &= e_t \\
var(e_t) &= f_t \\
\end{align*}
\] or, in the multivariate case

\[ 
var(e_t) = F_t = R_t R_t'
\]

\section{Algorithms}\label{algorithms}

State space models are convenient representations of time series to
derive multiple properties. Algorithms has been developed around them
for likelihood evaluation (and so, for parameters estimation), for
filtering/forecasting, for smoothing/missing values estimation, for
simulation\ldots{}

Multiple variations of those algorithms have been developed, to favor
either speed, memory use, robustness or accuracy.

The JD+ framework provides several family of those algorithms, mainly
based on approaches developed by Durbin-Koopman (DK), De Jong (AKF or
Augmented Kalman Filter), Kailath (Array filters) or Chandrasekhar
filters (CKMS or Chandrasekhar-Kailath-Morf-Sidhu). They will be
described in the chapter 3.

\section{Design}\label{design}

\chapter{Design}\label{design-1}

\section{Functional forms}\label{functional-forms}

State space forms and the related algorithms focus on the state vectors
and their (conditional) distribution, and on the relationships between
those vectors at different time points. For instance, using obvious
notations, we will consider:

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 2\tabcolsep) * \real{0.2571}}
  >{\raggedright\arraybackslash}p{(\linewidth - 2\tabcolsep) * \real{0.7429}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Operations
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Formulae
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Initialization & \(a_0 = a_{0\vert -1}\) \\
Prediction & \(a_{t \vert t}\rightarrow a_{t+1 \vert t}=a_{t+1}\) \\
Update & \(a_{t \vert t-1}=a_t \rightarrow a_{t \vert t}\) \\
Prediction error & \(e_t=y_t- \hat y_{t \vert t-1}\) \\
Smoothing & \(a_{t \vert n} \rightarrow a_{t-1\vert n}\) \\
\ldots{} & \ldots{} \\
\end{longtable}

The relationships considered above are usually expressed under the form
of matrices, which are often sparse. All things considered, matrices are
only a convenient way for describing linear transformations. By
replacing matrix computations (at least the most frequent and/or the
most expansive ones) with equivalent functions, it is possible to
achieve substantial performance gain.

This is the basic principle of the JD+ state space framework: every type
of model will have to provide a set of functions that allows an
efficient computation of the different algorithms considered in the
framework.

We shall clarify that point by an example. The prediction step is
defined by the equations:

\[ a_{t+1 \vert t}=T_t a_{t \vert t}, \quad P_{t+1 \vert t}=T_t P_{t \vert t} T_t'+V_t=T_t \left(T_t P_{t \vert t} \right)'+V_t \]

The transformation \(f_{T_t} (x)\) of an array (and by extension, of a
matrix) by means of the matrix \(T_t\) plays thus a central role in the
forecasting step:

\begin{itemize}
\tightlist
\item
  Apply \(f_{T_t}\) on \(a\)
\item
  Apply \(f_{T_t}\) on each column of \(P\) ; we get \(Q\)
\item
  Apply \(f_{T_t}\) on each row of \(Q\) ; we get \(O\)
\item
  Add \(V_t\) to \(O\)
\end{itemize}

In the case of (partially) time invariant systems, we will usually omit
in the documentation of the framework the subscript \(_t\) of the
different arrays/matrices/functions.

Using matrix computation, the transformation \(f_{T_t}\) needs
\(r \times r\) multiplications, where \(r\) is the length of the state
array. In many cases, the functional form will involve at most \(r\)
multiplications and much less memory traffic.

For instance, in some SSF forms of ARIMA models (see Gomez-Maravall,
1994), the function \(y=f_T(x)\) is defined by

\[ y_i = \begin{cases} x_{i+1} & 0 \le i < r-1  \\ -\sum_{k=1}^p {\varphi_k x_{r-k} } & i =r-1\end{cases} \]

which corresponds to the transformation matrix

\[
\begin{pmatrix}
0 & 1 & 0 & \cdots & 0 \\
0 & 0 & 1 & \cdots & 0 \\
\vdots & \vdots & \ddots & \ddots & 0 \\
0 & \cdots & \cdots & 0 & 1 \\
0 & \cdots & -\phi_p & - \cdots & -\phi_1
\end{pmatrix}
\]

It involves only \(p\) (order of the autoregressive polynomial)
multiplications.

As soon as direct matrix computations are avoided, the matrices of the
system themselves don't usually need to be created. That leads to
another substantial gain in time and in memory space/traffic, especially
in the case of time dependent systems.

\chapter{SSF structure}\label{ssf-structure}

\chapter{Java design}\label{java-design}

\part{SSF blocks}

\chapter{Atomic blocks}\label{atomic-blocks}

\section{ARMA model}\label{arma-model}

\subsection{Introduction}\label{introduction-1}

The ARMA process is defined by

\[\Phi\left(B\right)y_t=\Theta\left(B\right)\epsilon_t \]

where:

\[\Phi\left(B\right)=1+\varphi_1 B + \cdots + \varphi_p B^p \]

\[\Theta\left(B\right)=1+\theta_1 B + \cdots + \theta_q B^q \]

are the auto-regressive and the moving average polynomials.

The MA representation of the model is
\(y_t=\sum_{i=0}^\infty {\psi_i \epsilon_{t-i}}\). Let \(\gamma_i\) be
the autocovariances of the model. We also define:
\(r=\max\left(p, q+1\right), \quad s=r-1\).

The state-space model can be written in different ways. JD+ provides two
of them

\subsection{Representation I}\label{representation-i}

\subsubsection{State vector}\label{state-vector}

\[ \alpha_t= \begin{pmatrix} y_t \\ y_{t+1|t} \\ \vdots \\ y_{t+s|t} \end{pmatrix}\]

where \(y_{t+i|t}\) is the orthogonal projection of \(y_{t+i}\) on the
subspace generated by \({y\left(s\right):s \leq t}\).Thus, it is the
forecast function with respect to the semi-infinite sample. We also have
that \(y_{t+i|t} = \sum_{j=i}^\infty {\psi_j \epsilon_{t+i-j}}\)

\subsubsection{Dynamics}\label{dynamics}

\[ T_t = \begin{pmatrix}0 &1 & 0 & \cdots & 0  \\0& 0 & 1 & \cdots & 0\\ \vdots & \vdots & \vdots & \ddots & \vdots\\ 0 & 0 & 0 & \cdots & 1\\
-\varphi_r & \cdots  & \cdots & \cdots &-\varphi_1 \end{pmatrix}\]

with \(\varphi_{j}=0\) for \(j>p\)

\[ S_t = \begin{pmatrix}1 \\ \psi_1 \\ \vdots\\ \psi_s \end{pmatrix} \]

\[ V_t = S S' \]

\subsubsection{Default measurement}\label{default-measurement}

\[ Z_t = \begin{pmatrix} 1 & 0 & \cdots & 0\end{pmatrix}\]

\[ h_t = 0 \]

\subsubsection{Initialization}\label{initialization}

\[ \alpha_{-1} = \begin{pmatrix}0 \\ 0 \\ \vdots\\ 0 \end{pmatrix} \]

\[ P_{*} = \Omega \]

\(\Omega\) is the unconditional covariance of the state array; it can be
easily derived using the MA representation. We have:

\[ \Omega\left(i,0\right) = \gamma_i \]

\[ \Omega\left(i,j\right) = \Omega\left(i-1,j-1\right)-\psi_i \psi_j \]

\subsection{Representation II}\label{representation-ii}

We provide below an alternative state space representation of ARMA
models

\subsubsection{State vector:}\label{state-vector-1}

\[ \alpha_t= \begin{pmatrix} y_t \\ 
-\varphi_2 y_{t-1}-\dots-\varphi_r y_{t-r+1}+\theta_1 \epsilon_t+ \dots + \theta_{r-1} \epsilon_{t-r+2}\\ -\varphi_3 y_{t-1}-\dots-\varphi_r y_{t-r+2}+\theta_2 \epsilon_t+ \dots + \theta_{r-1} \epsilon_{t-r+3}\\ \vdots \\ 
-\varphi_r y_{t-1}+\theta_{r-1} \epsilon_{t}
\end{pmatrix}\]

\subsubsection{Dynamics}\label{dynamics-1}

\[ T_t = \begin{pmatrix}-\varphi_1 &1 & 0 & \cdots & 0  \\
-\varphi_2& 0 & 1 & \cdots & 0\\ 
\vdots & \vdots & \vdots & \ddots & \vdots\\ 
\vdots & 0 & 0 & \cdots & 1\\
-\varphi_r & 0  & \cdots & \cdots &0 \end{pmatrix}\]

\[ S_t = \begin{pmatrix}1 \\ \theta_1 \\ \vdots\\ \theta_r \end{pmatrix} \]

\[ V_t = S S' \]

\subsubsection{Default measurement}\label{default-measurement-1}

\[ Z_t = \begin{pmatrix} 1 & 0 & \cdots & 0\end{pmatrix}\]

\[ h_t = 0 \]

\subsubsection{Initialization}\label{initialization-1}

\[ \alpha_{-1} = \begin{pmatrix}0 \\ 0 \\ \vdots\\ 0 \end{pmatrix} \]

\[ P_{*} = \Omega \]

\(\Omega\) is the unconditional covariance of the state array; it can be
easily derived using the MA representation. We have:

\[ \Omega\left(i,0\right) = \gamma_i \]

\[ \Omega\left(i,j\right) = \Omega\left(i-1,j-1\right)-\psi_i \psi_j \]
\#\#\# Implementation

\section{Time dependent ARMA model}\label{time-dependent-arma-model}

\subsection{Introduction}\label{introduction-2}

The time dependent ARMA process (TDARMA) is defined by

\[ \Phi_t\left(B\right)y_t=\Theta_t\left(B\right)\epsilon_t \]

where:

\[\Phi_t\left(B\right)=1+\varphi_{t,1} B + \cdots + \varphi_{t,p} B^p \]

\[\Theta_t\left(B\right)=1+\theta_{t,1} B + \cdots + \theta_{t,q} B^q \]

are the time dependent auto-regressive and the moving average
polynomials.

\subsection{State vector:}\label{state-vector-2}

\[ \alpha_t= \begin{pmatrix} y_t \\ 
-\varphi_{t+1,2} y_{t-1}-\dots-\varphi_{t+1,r} y_{t-r+1}+\theta_{t+1,1} \epsilon_t+ \dots + \theta_{t+1, r-1} \epsilon_{t-r+2}\\ -\varphi_{t+2,3} y_{t-1}-\dots-\varphi_{t+2,r} y_{t-r+2}+\theta_{t+2,2 }\epsilon_t+ \dots + \theta_{t+2,r-1} \epsilon_{t-r+3}\\ \vdots \\ 
-\varphi_{t+r-1,r} y_{t-1}+\theta_{t+r-1,r-1} \epsilon_{t}
\end{pmatrix}\]

\subsection{Dynamics}\label{dynamics-2}

\[ \alpha_{t+1} = T_t \alpha_t + \begin{pmatrix} 1 \\ \theta_{t+2,1} \\ \theta_{t+3,2} \\ \vdots \\  \theta_{t+r, r-1} \\ \end{pmatrix} \epsilon_{t+1} \]

\[ T_t = \begin{pmatrix}-\varphi_{t+1,1} &1 & 0 & \cdots & 0  \\
-\varphi_{t+2,2} & 0 & 1 & \cdots & 0\\ 
\vdots & \vdots & \vdots & \ddots & \vdots\\ 
\vdots & 0 & 0 & \cdots & 1\\
-\varphi_{t+r,r} & 0  & \cdots & \cdots &0 \end{pmatrix}\]

\[ S_t = \begin{pmatrix}\sigma_{t+1}\\ \sigma_{t+1} \theta_{t+2,1} \\ \vdots\\ \sigma_{t+1} \theta_{t+r,r-1} \end{pmatrix} \]

\[ V_t = S S' \]

\subsection{Default measurement}\label{default-measurement-2}

\[ Z_t = \begin{pmatrix} 1 & 0 & \cdots & 0\end{pmatrix}\]

\[ h_t = 0 \]

\subsection{Initialization}\label{initialization-2}

\[ \alpha_0= \begin{pmatrix} y_0 \\ 
-\varphi_{1,2} y_{-1}-\dots-\varphi_{1,r} y_{-r+1}+\theta_{1,1} \epsilon_0+ \dots + \theta_{1, r-1} \epsilon_{-r+2}\\ -\varphi_{2,3} y_{-1}-\dots-\varphi_{2,r} y_{-r+2}+\theta_{2,2 }\epsilon_0+ \dots + \theta_{2,r-1} \epsilon_{-r+3}\\ \vdots \\ 
-\varphi_{r-1,r} y_{-1}+\theta_{r-1,r-1} \epsilon_{0}
\end{pmatrix}\]

The covariance of \(\alpha_0\) can be easily computed, taking into
account that \(y_0, \dots,y_{r-1}\) and
\(\epsilon_0, \dots, \epsilon_{r-1}\) are generated by the same
processes and that their variances/covariances are:

\[ cov(y_i, y_j)=\gamma(i-j)\]

\[ cov(y_i, \epsilon_j) = \psi_{j-i}\sigma^2, j\le i\]

\[ cov(\epsilon_i, \epsilon_j)=\sigma^2, i=j \]

\section{ARIMA model}\label{arima-model}

\subsection{Introduction}\label{introduction-3}

We consider that the auto-regressive polynomial contains some unit
roots. It can be factorized in a stationary polynomial (defined by the
roots outside the unit circle) and in a non-stationary polynomial
(defined by the roots on the unit circle), which is notated:

\[ \Delta\left(B\right) = 1 + \delta_1 B + \cdots + \delta_d B^d \]

The state space form of an ARIMA model is similar to the state space
form of an ARMA model, except for its initialization.

\subsection{Initialization}\label{initialization-3}

The initial conditions can be written as follows:

\[ \alpha_{-1} = \begin{pmatrix}1 \\ 0 \\ \vdots\\ 0 \end{pmatrix} \]

\[ P_{*} = \Sigma \Omega \Sigma' \]

\[ B = \Lambda \]

\[ P_{\infty}= \Lambda \Lambda' \]

\(\Omega\) is the unconditional covariance of the state array of the
stationary model.

\[ \Sigma = \begin{pmatrix} 1 & 0 & \cdots & 0 \\ \lambda_1 & 1 & \cdots & 0 \\ \lambda_2 & \lambda_1 & \cdots & \vdots\\ \vdots & \vdots & \vdots & \vdots \\ \lambda_{r-1} & \lambda_{r-2} & \cdots & \lambda_{1} \end{pmatrix} \]

where \(\lambda_{i}\) are generated by
\(\frac{1}{\Delta\left(B\right)}\)

\(\Lambda\) is a r x d matrix; its first d rows form an identity matrix;
other cells are defined by the recursive relationship:

\[\Lambda \left(i,j\right) = -\sum_{k=1}^d {\delta_k \Lambda \left(i-k,j\right)}\]

\section{Noise}\label{noise}

\subsection{State vector}\label{state-vector-3}

\subsection{Dynamics}\label{dynamics-3}

\subsection{Default measurement}\label{default-measurement-3}

\subsection{Initialization}\label{initialization-4}

\subsection{Implementation}\label{implementation}

\section{Local level}\label{local-level}

\subsubsection{Description}\label{description}

The local level block describes a random walk component.

\[
\begin{align*}
l_{t+1} &= l_t + \epsilon_t \\
\epsilon_t &\sim N(0, \sigma^2_l)
\end{align*}
\]

\subsection{State vector}\label{state-vector-4}

The state block is \(\alpha_t=\begin{pmatrix} l_t  \end{pmatrix}\)

\subsection{Dynamics}\label{dynamics-4}

\[
\begin{align*}
T_t &= \begin{pmatrix} 1 \end{pmatrix} \\
V_t &= \begin{pmatrix}\sigma^2_l \end{pmatrix}\\
S_t &= \begin{pmatrix}\sigma_l \end{pmatrix}
\end{align*}
\]

\subsection{Default measurement}\label{default-measurement-4}

\[
\begin{align*}
Z_t &= \begin{pmatrix} 1 \end{pmatrix} \end{align*} 
\]

\subsection{Diffuse initialization}\label{diffuse-initialization}

\[
\begin{align*}
a_0 &= \begin{pmatrix} 0 \end{pmatrix} \\
P_* &= \begin{pmatrix} 0 \end{pmatrix}\\
B &= \begin{pmatrix} 1 \end{pmatrix} \\
P_\infty &= \begin{pmatrix} 1 \end{pmatrix}
\end{align*}
\]

\subsection{Fixed initialization}\label{fixed-initialization}

\[
\begin{align*}
a_0 &= \begin{pmatrix} l_0 \end{pmatrix} \\
P_* &= \begin{pmatrix} \sigma_{l} \end{pmatrix}\\
B &= \begin{pmatrix} 0 \end{pmatrix} \\
P_\infty &= \begin{pmatrix} 0 \end{pmatrix}
\end{align*}
\]

\subsection{Implementation}\label{implementation-1}

See \texttt{jdplus.toolkit.base.core.ssf.sts.LocalLevel}

\section{Local linear trend}\label{local-linear-trend}

\subsection{Description}\label{description-1}

The local linear trend block describes the following trend component:

\[
\begin{align*}
l_{t+1} &= l_t + n_t +  \epsilon_t \\
n_{t+1} &= n_t + \mu_t \\
\epsilon_t &\sim N(0, \sigma^2_l) \\
\mu_t &\sim N(0, \sigma^2_n)
\end{align*}
\]

\subsection{State vector}\label{state-vector-5}

The state block is

\[\alpha_t=
\begin{pmatrix} 
l_t \\ 
n_t  
\end{pmatrix}
\]\\

\subsection{Dynamics}\label{dynamics-5}

\[
\begin{align*}
T_t &= \begin{pmatrix} 1 && 1 \\ 0 && 1 \end{pmatrix} \\
V_t &= \begin{pmatrix} \sigma^2_l && 0 \\ 0 && \sigma^2_n \end{pmatrix} \\
S_t &= \begin{pmatrix} \sigma_l && 0 \\ 0 && \sigma_n \end{pmatrix} 
\end{align*}
\]

\subsection{Default measurement}\label{default-measurement-5}

\[
\begin{align*}
Z_t &= \begin{pmatrix} 1 & 0 \end{pmatrix} 
\end{align*} 
\]

\subsection{Initialization}\label{initialization-5}

\[
\begin{align*}
a_0 &= \begin{pmatrix} 0 \\ 0  \end{pmatrix} \\
P_* &=  \begin{pmatrix} 0 && 0 \\ 0 && 0 \end{pmatrix}\\
B &= \begin{pmatrix} 1 && 0 \\ 0 && 1 \end{pmatrix} \\
P_{\infty} &= \begin{pmatrix} 1 && 0 \\ 0 && 1 \end{pmatrix}
\end{align*}
\]

\subsection{Parameters}\label{parameters}

\[ \sigma^2_l \ge 0,\quad \sigma^2_n \ge 0 \]

\subsection{Implementation}\label{implementation-2}

See \texttt{jdplus.toolkit.base.core.ssf.sts.LocalLinearTrend}

\section{Seasonal component}\label{seasonal-component}

\subsection{Introduction}\label{introduction-4}

The seasonal component provided in JD+ is based on Proietti{[}2000{]}

We consider that it is composed of latent variables corresponding to
each period, which follow a multivariate random walk. We also impose
that the sum of the latent variables is 0 at each point time.

\subsection{State vector}\label{state-vector-6}

For a periodicity of \(s\), we define the state space form of the
seasonal component as follows (West-Harrison form):

We consider first the state vector \(\tilde \gamma_t\) such that
\(\tilde \gamma_{it}\) is the (unobserved) seasonal component of season
\(i\) observed at time \(t\).

Using that definition would imply a time varying measurement error. To
avoid it, we use the state vector \(\gamma_t\) which is a rotation of
\(\tilde \gamma_t\) such that the first component corresponds to the
current period. More precisely,
\(\gamma_{it} =\tilde\gamma_{\left( \left(t-i\right) \mod s \right), t}\).

Moreover, the constraint that the sum of the latent variables is 0
allows us to consider only \(s-1\) periods (the missing one being the
opposite of the sum of all the other periods).

\subsection{Dynamics}\label{dynamics-6}

\[ T_t = \begin{pmatrix} 0 & 1 & \cdots & 0 \\ \vdots & \ddots & \ddots & \vdots \\ 0 &  \cdots & 0 & 1  \\ -1 & -1 & \cdots & -1\end{pmatrix}\]

The covariance matrix of the innovations can take various form. Using
the terminology of Proietti, we consider:

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Model
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
V
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
S
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Dummy &
\(\begin{pmatrix} 0 &  \cdots & 0 \\ \vdots & & \vdots\\ 0 &\cdots &\sigma_{seas}^2\end{pmatrix}\)
& \(\begin{pmatrix}0 \\ \vdots \\ \sigma_{seas}\end{pmatrix}\) \\
Crude &
\(\sigma_{seas}^2\begin{pmatrix} 1 &  \cdots & 1 \\ \vdots & & \vdots\\ 1 &\cdots & 1\end{pmatrix}\)
& \(\sigma_{seas}\begin{pmatrix}1 \\ \vdots \\ 1\end{pmatrix}\) \\
HarrisonStevens & \(\sigma\_{seas}\^2\) & \\
Trigonometric & & \\
\end{longtable}

\subsection{Measurement}\label{measurement}

\[ Z_t = \begin{pmatrix} 1 & 0 & \cdots & 0\end{pmatrix}\]

\subsection{Bibliography}\label{bibliography}

Proietti T. (2000), ``Comparing seasonal components for structural time
series models'', International Journal of Forecasting, 16, 2, 247-260.

\section{Cyclical component}\label{cyclical-component}

\subsection{Introduction}\label{introduction-5}

A Cycle component is a special case of an AR2 component, which contains
two complex conjugate roots.

More precisely, it is defined by a period \(\lambda\) and a dumping
factor \(\rho\). If \(\gamma = \frac{2\pi}{\lambda}\), we have that

\[   \begin{pmatrix} y_{t+1} \\ y^*_{t+1}\end {pmatrix} = \begin{pmatrix} \rho \cos{\gamma} &&   \rho \sin{\gamma}\\  -\rho \sin{\gamma}  && \rho \cos{\gamma}\end {pmatrix}\begin{pmatrix} y_t \\ y_t^*\end {pmatrix}+\begin{pmatrix} \epsilon_t \\ \epsilon_t^*\end {pmatrix} \]

Using those notations, the state-space model can be written as follows :

\subsection{State vector:}\label{state-vector-7}

\[ \alpha_t= \begin{pmatrix} y_t \\ y_t^* \end{pmatrix}\]

\subsection{Dynamics}\label{dynamics-7}

\[ T_t = \begin{pmatrix} \rho \cos{\gamma} &&   \rho \sin{\gamma}\\  -\rho \sin{\gamma}  && \rho \cos{\gamma}\end {pmatrix}\]

\[ S_t = \sigma_c \begin{pmatrix} 1 && 0\\  0  && 1\end {pmatrix} \]

\[ V_t = \sigma_c^2\begin{pmatrix} 1 && 0\\  0  && 1\end {pmatrix} \]

\subsection{Measurement}\label{measurement-1}

\[ Z_t = \begin{pmatrix} 1 & 0 \end{pmatrix}\]

\[ h_t = 0 \]

\subsection{Initialization}\label{initialization-6}

\[
\begin{align*}
a_0 &= \begin{pmatrix}0 \\ 0 \end{pmatrix} \\
P_{*} &=  \sigma_c^2\begin{pmatrix} \frac{1}{1-\rho^2} && 0\\  0  && \frac{1}{1-\rho^2}\end {pmatrix}
\end{align*}
\]

\chapter{Derived blocks}\label{derived-blocks}

\part{Algorithms}

\chapter{Ordinary filters}\label{ordinary-filters}

\section{Filters}\label{filters}

\subsection{Univariate model}\label{univariate-model}

\subsubsection{Update step t}\label{update-step-t}

\[
\begin{align*}
e_t &= y_t - Z_t a_t \\
M_t &= P_t Z_t' \\
f_t &= Z_t P_t Z_t' +h_t = M_tZ_t' + h_t \\
a_{t|t} &= a_t + M_t f_t^{-1}e_t \\
P_{t|t}&= P_t - M_t f_t^{-1} M_t'
\end{align*}
\]

\subsubsection{Forecast step t}\label{forecast-step-t}

\[
\begin{align*}
a_{t+1} &= T_t a_{t|t} \\
P_{t+1} &= T_t P_{t|t} T_t' + V_t
\end{align*}
\]

\subsubsection{Compact form}\label{compact-form}

The two steps can be easily combined to give a more compact formulation.

\[
\begin{align*}
e_t &= y_t - Z_t a_t \\
f_t &= Z_t P_t Z_t' + h_t \\
K_t &= T_t P_t Z_t' f_t^{-1} \\
a_{t+1} &= T_t a_t + K_t e_t \\
P_{t+1} &= T_t P_t T_t' - K_t f_t K_t' + V_t
\end{align*}
\]

However, the current implementation uses explicitly the two steps form,
to be able to store either \(a_t\) and \(a_{t|t}\).

\subsection{Multivariate model}\label{multivariate-model}

In the multi-variate case, we use a slightly different (but strictly
equivalent) implementation:

\subsubsection{Update step t}\label{update-step-t-1}

\[
\begin{align*}
e_t &= y_t - Z_t a_t \\
F_t &= Z_t P_t Z_t' + H_t = R_t R_t' \quad(Cholesky) \\
\tilde{M_t} &= P_t Z_t' {R_t'}^{-1} \Leftrightarrow \tilde{M_t} R_t' = P_t Z_t' \\
u_t &= R_t^{-1} e_t \Leftrightarrow R_t u_t = e_t \\
a_{t|t} &= a_t + \tilde{M_t} u_t \\
P_{t|t} &= P_t - \tilde{M_t}\tilde{M_t}'
\end{align*}
\]

\subsubsection{Forecast step t}\label{forecast-step-t-1}

\[
\begin{align*}
a_{t+1} &= T_t a_{t|t} \\
P_{t+1} &= T_t P_{t|t} T_t' + V_t
\end{align*}
\]

\subsubsection{Compact form}\label{compact-form-1}

\[
\begin{align*}
e_t &= y_t - Z_t a_t \\
F_t &= Z_t P_t Z_t' + H_t = R_t R_t' \quad(Cholesky) \\
\tilde{K_t} &=  T_t P_t Z_t' {R_t'}^{-1} \Leftrightarrow \tilde{K_t} R_t' = T_t P_t Z_t' \\
u_t &= R_t^{-1} e_t \Leftrightarrow R_t u_t = e_t \\
a_{t+1} &= T_t a_t + \tilde{K_t} u_t \\
P_{t+1} &= T_t P_{t} T_t' + V_t - \tilde{K_t}\tilde{K_t}' 
\end{align*}
\]

\subsubsection{Special cases}\label{special-cases}

This implementation is robust (the covariance matrices are symmetric by
construction) and makes the computation of the likelihood easy. It also
provides a straightforward solution for singular covariance matrices and
for missing values.

When some observations are missing, we just remove from \(Z_t\) the
corresponding equations.

When the covariance matrix \(F_t\) is singular, some columns of \(R_t\)
are equal to 0. Those columns correspond to ``redundant'' equations. The
corresponding errors \(u_t\) and the corresponding columns in the gain
matrix are undefined (set to 0). The over-determined system
\(R_t u_t = e_t\) imposes some restrictions on the observations, which
can be easily checked (the system must be solvable).

When the measurement errors are diagonal, it should be noted that the
solution based on the Cholesky decomposition is identical to the
so-called univariate treatment of multi-variate models.

The following elemnts are saved in objects of the class
\texttt{MultivariateFilteringResults}

\(a_t, P_t, U_t, R_t, \tilde{K_t}\)

\subsection{Implementation}\label{implementation-3}

The ordinary filter is implemented in the classes
\texttt{jdplus.toolkit.base.core.ssf.univariate.OrdinaryFilter} and
\texttt{jdplus.toolkit.base.core.ssf.multivariate.MultivariateOrdinaryFilter}.

\section{Smoothers}\label{smoothers}

\subsection{Univariate model}\label{univariate-model-1}

\subsubsection{Notations}\label{notations}

We describe hereafter the recursions of the ordinary smoother. The
equations related to the variances can be omitted.

\(a_t, P_t, M_t \left(=P_t Z_t' \right), f_t, e_t\) are quantities
obtained in the filtering process. \(r_t\) are auxiliary row-matrices,
while \(N_t\) are auxiliary square matrices. \(r_t,N_t\) are set to 0 at
the beginning of the smoothing process.

The following notations are used:

\[
\begin{align*}
\tilde a_t &=E\left(\alpha_{t} \vert y_0 \cdots y_{n-1}\right) \\
\tilde P_t &=var\left(\alpha_{t} \vert y_0 \cdots y_{n-1}\right) \\
\tilde e_t &=E\left(\epsilon_{t} \vert y_0 \cdots y_{n-1}\right)
\end{align*}
\]

\subsubsection{\texorpdfstring{Recursion for observed \(y_t\) and
\(f_t > 0\)}{Recursion for observed y\_t and f\_t \textgreater{} 0}}\label{recursion-for-observed-y_t-and-f_t-0}

\[
\begin{align*}
\text{recall: }K_t &= T_t P_t Z_t' f_t^{-1} = T_t M_t f_t^{-1}\\
\tilde e_t &= e_t / f_t - r_t K_t \\
var(\tilde e_t) &= 1/f_t + K_t' N_t K_t \\
r_{t-1} &= \tilde e_t Z_t + r_t T_t \\
L_t &= T_t - K_t Z_t \\
N_{t-1} &= Z_t' Z_t / f_t + L_t' N_t L_t 
\end{align*}
\]

\subsubsection{\texorpdfstring{Recursion for missing \(y_t\) or
\(f_t = 0\)}{Recursion for missing y\_t or f\_t = 0}}\label{recursion-for-missing-y_t-or-f_t-0}

\[
\begin{align*}
 r_{t-1} &= r_t T_t \\
N_{t-1} &= T_t' N_t T_t
\end{align*}
\]

\subsubsection{Smoothed states}\label{smoothed-states}

The smoothed states are then obtained by

\[
\begin{align*}
\tilde a_t' &= a_t' + r_t P_t \\
\tilde P_t &= P_t + P_t N_t P_t
\end{align*}
\]

\subsubsection{Smoothed disturbances}\label{smoothed-disturbances}

The smoothed disturbances, which can be useful for diagnostics, are
computed by

\[
\begin{align*}
\tilde u_t &= r_t S_t \\
var\left(\tilde u_t \right) &= V_t-S_t' N_t S_t
\end{align*}
\]

\subsubsection{Implementation details}\label{implementation-details}

The ordinary smoother for univariate models is implemented in the class
\texttt{jdplus.toolkit.base.core.ssf.univariate.OrdinarySmoother}

Most equations are straightforward and uses functional forms. The most
expensive part is the computation of \(\tilde P_t\), for which we
process as follows:

\[
\begin{align*}
 A &=  xl(xl(N_t)) \\
 N_{t-1} &= Z_t' Z_t / f_t + A
\end{align*}
\]

where the operator \(xl(x)=x(T_t - K_t Z_T)\) is computed following the
fast equations:

\[
\begin{align*}
q &= x T_t \\
w &= q M_t \\
xl(x) &= q - w f_t^{-1} Z_t
\end{align*}
\] That transformation, which doesn't imply explicit matrix
computations, is applied successively on the columns and on the rows of
\(N_t\)

\subsection{Multivariate model}\label{multivariate-model-1}

\(a_t, P_t, \tilde M_t \left(\text{such that }M_t R'_t = P_t Z_t' \right), R_t, u_t\left(\text{such that } u_t R'_t = e_t \right)\)
are quantities obtained in the filtering process. \(R_t\) is the
Cholesky factor of \(F_t\) (\(R_t R'_t = F_t\)), It can contain
rows/columns equal to 0 when \(F_t\) is non invertible.

The previous algorithm becomes:

\subsubsection{\texorpdfstring{Recursion for observed
\(y_t\)}{Recursion for observed y\_t}}\label{recursion-for-observed-y_t}

\[
\begin{align*}
\tilde K_t &= T_t \tilde M_t \\
\tilde e_t R_t &= (u_t - r_t \tilde K_t) \\
R'_t var(\tilde e_t) R_t &= I+ K_t' N_t K_t \\
r_{t-1} &= \tilde e_t Z_t + r_t T_t \\
L_t &= T_t - K_t Z_t \\
Z_t &= R_t W_t \\
N_{t-1} &= W_t' W_t + L_t' N_t L_t
\end{align*}
\] \#\#\#\# Recursion for missing \(y_t\) \$

\[
\begin{align*}
 r_{t-1} &= r_t T_t \\
N_{t-1} &= T_t' N_t T_t
\end{align*}
\]

\chapter{Chandrasekhar filters}\label{chandrasekhar-filters}

\chapter{Array filters}\label{array-filters}

\chapter{Likelihood}\label{likelihood}

We write \(l\) the log of the likelihood

\section{Stationary models}\label{stationary-models}

When the initial state of the model is completely specified, the Kalman
filter provides the exact likelihood in a trivial way, using the
prediction error decomposition:

\[ p(y_0, \ldots, y_{n-1}) = p(y_0) p(y_1\vert y_0) \ldots p(y_{n-1}\vert y_0 \ldots, y_{n-2})\]

Using the notations defined for the ordinary filter, we have:

\subsection{univariate case:}\label{univariate-case}

\[
l(y;\theta, \sigma^2)=-1/2\left(n \log(2 \pi) + n \log(\sigma^2)  +  \frac{1}{\sigma^2}\sum_{i=0}^{i<n} e_i^2/f_i + \sum_{i=0}^{i<n} {\log \vert f_i \vert }\right)
\]

By concentrating \(\sigma^2\) out of the likelihood and by writing
\(RSS=\sum_{i=1}^n e_i^2/f_i\), we obtain :

\[
\begin{align*}
\hat{\sigma}^2 &= \frac{RSS}{n} \\
l_c(y; \theta)&=-1/2\left(n \log(2 \pi) + n - n \log(n) + n \log(RSS) + \sum_{i=0}^{i<n} {\log \vert f_i \vert }\right)
\end{align*}
\]

\subsection{multi-variate case:}\label{multi-variate-case}

if we write \(N=\sum_{i=0}^n{T_i}\) where \(T_i\) is the number of
observations for the period \(i\) and \(RSS=\sum_{i=1}^n u_i'u_i\), we
get:

\[
\begin{align*}
l(y; \theta, \sigma^2) &=-1/2\left(N \log(2 \pi) + N \log(\sigma^2)  +  \frac{1}{\sigma^2}\sum_{i=0}^{i<n} u_i'u_i + 2\sum_{i=0}^{i<n} {\log \vert R_i \vert }\right) \\
\hat{\sigma}^2 &= \frac{RSS}{N} \\
l_c(y; \theta) &=-1/2\left(N \log(2 \pi) + N  -N \log(N) + N\log(RSS) + 2\sum_{i=0}^{i<n} {\log \vert R_i \vert }\right) 
\end{align*}
\]

\section{Diffuse models}\label{diffuse-models}

\subsection{State space models as linear regression
models}\label{state-space-models-as-linear-regression-models}

The general linear gaussian state space model can be transformed in the
regression model

\[z = y - c = X \delta + \mu  \] where

\[c_t= Z_t\left[ \left(\prod_{i=0}^{i<t}{T_i}\right)\right]b \quad X_t = Z_t\left[ \left(\prod_{i=0}^{i<t}{T_i}\right)\right]B \]

and the covariance matrix \(\sigma^2\Omega\) of \(\mu\) depends on the
structure of the state space model.

As discussed in the paragraph \ldots, the Kalman filter defines the
linear transformation: \(z \rightarrow e_z/\sqrt{f}\). We write \(z_L\)
the result of that transformation.

We present below the univariate case. The multi-variate models are
handled in a similar way.

\subsection{Profile likelihood}\label{profile-likelihood}

When \(\delta\) is considered as fixed unknown, the log-likelihood is:

\[
l(y;\theta, \delta, \sigma^2)=-1/2\left(n \log(2 \pi) + n \log(\sigma^2)  +  \frac{1}{\sigma^2}\sum_{i=0}^{i<n} (y_i-c_i-X_i \delta)_L^2 + \sum_{i=0}^{i<n} {\log \vert f_i \vert }\right)
\]

\((y-c-X \delta)_L = (y-c)_L - X_L \delta\) and we can concentrate
\(\delta\) and \(\sigma^2\) out of the likelihood.

\[ \begin{align*}
\hat{\delta} &= (X_L'X_L)^{-1}X'_L(y-c)_L \\
\hat{e}_L &= (y-c)_L - X_L \hat{\delta} \\
RSS &= \hat{e}_L' \hat{e}_L \\
\hat{\sigma}^2 &= \frac{RSS}{n} \\
l_c(y; \theta)&=-1/2\left(n \log(2 \pi) + n - n \log(n) + n \log(RSS) + \sum_{i=0}^{i<n} {\log \vert f_i \vert }\right)
\end{align*} 
\]

\subsection{Diffuse likelihood}\label{diffuse-likelihood}

When \(\delta\) is random ( \(\delta \sim N(0, \kappa \sigma^2 I)\) ),
the log-likelihood is given by:

\[
l(y; \theta, \sigma^2) = l(y \vert \delta; \theta, \sigma^2) + l(\delta; \theta, \sigma^2) - l(\delta \vert y; \theta , \sigma^2)  
\]

If d is the dimension of \(\delta\), we have that:

\[
\begin{align*}
l(\delta; \theta, \sigma^2) &= -1/2 \left(d \log(2 \pi) + d \log(\sigma^2) + d \log(\kappa) + \frac{1}{\kappa \sigma^2}{\delta'\delta}  \right) \\
l(y \vert \delta, \theta, \sigma^2) &= -1/2\left(n \log(2 \pi) + n \log(\sigma^2)  +  \frac{1}{\sigma^2}(y -c - X \delta)_L'(y -c - X \delta)_L + \sum_{i=0}^{i<n} {\log \vert f_i \vert }\right) \\ 
E(\delta \vert y; \theta, \sigma^2) &= (\frac{1}{\kappa}I + {X_L}' X_L)^{-1}{X_L}'(y-c)_L = \Delta_\kappa^{-1}{X_L}'y^c_L \\
var(\delta \vert; \theta, \sigma^2) &= \sigma^2 (\frac{1}{\kappa}I + {X_L}' X_L)^{-1} = \sigma^2  \Delta_\kappa^{-1}\\
l(\delta \vert y; \theta , \sigma^2) &=  -1/2 \left(d \log(2 \pi) + d \log(\sigma^2) - \log \vert \Delta_\kappa \vert + \frac{1}{\sigma^2}\delta' \Delta_\kappa \delta + \frac{1}{\sigma^2}{y^c_L}'X_L \Delta_\kappa^{-1}{X_L}'y^c_L -\frac{2}{\sigma^2} {y^c_L}'X_L \delta  \right)
\end{align*}
\]

Following Francke et al, we define the diffuse log-likelihood as:

\[
l_\infty(y; \theta, \sigma^2) = \lim_{\kappa \rightarrow \infty}{l(y;\theta, \sigma^2) + \frac{1}{2}\log(\vert 2 \pi \sigma^2 \kappa I)  \vert} 
\]

so that, writing \(m=n-d\)

\[
l_\infty(y; \theta, \sigma^2) = -1/2 \left( m \log(2 \pi) + m - m \log m + m \log RSS + \log \vert {X_L}' X_L \vert + \sum_{i=0}^{i<n} {\log \vert f_i \vert }\right)
\]

\subsection{Marginal likelihood}\label{marginal-likelihood}

The marginal likelihood is the likelihood of \(y^*=Ay\), which doesn't
depend on \(\delta\). The transformation A has dimensions \(n\times m\)
and is such that \(A'X=0\)

\[
\begin{align*}
l(y^*; \theta, \sigma^2) &= -1/2\left(m \log 2\pi + m \log \sigma^2 + \log({A_L}'A_L)  + \frac{1}{\sigma^2}{y^c}'A({A_L}'A_L)^{-1}A'y^c\right) \\
l(y^*; \theta, \sigma^2) &= -1/2\left(m \log 2\pi + m \log \sigma^2 + \log \vert {X_L}'X_L \vert +\frac{1}{\sigma^2} RSS  - \log \vert X'X \vert  +  \sum_{i=0}^{i<n} {\log \vert f_i \vert }\right) \\
l(y^*; \theta, \sigma^2) &= -1/2\left(m \log 2\pi + m - m \log m + m \log RSS + \log \vert {X_L}' X_L \vert - \log \vert X'X \vert + \sum_{i=0}^{i<n} {\log \vert f_i \vert } \right)
\end{align*}
\]

\subsection{Comments}\label{comments}

\begin{itemize}
\tightlist
\item
  On occasion, we want to consider some regression parameters as fixed
  unknown and others as diffuse. That's the case with regression
  variables and non-stationary noises (for instance in REGARIMA models).
  In that case, we will include the diffuse correction only for the
  regression components related to the diffuse initialization.
\item
  Other algorithms are available for the computation of the likelihood
  of diffuse models. They usually try to exploit the special structure
  of the regression variables. They are often significantly faster, but
  less stable. The variant of the current definitions based on a QR
  decomposition fo the computation of the RSS presents a robust solution
  and should be the preferred one as far as robustness is favored.
\item
  the ML estimator of \(\sigma^2\) is biased. We usually will correct it
  (using \(m=n - d\)) in various tests (not in our definition of the
  likelihood).
\end{itemize}

\subsection{Implementation}\label{implementation-4}

The univariate version of the transformation of the diffuse elements in
regression variables is implemented by the method
\texttt{jdplus.toolkit.base.core.ssf.diffuseEffects} and the
transformation of \(a_0\) by
\texttt{jdplus.toolkit.base.core.ssf.initialEffect}

The different likelihood functions are implemented in
\texttt{jdplus.toolkit.base.core.ssf.akf.QRFilter}. It should be noted
that the RSS are computed in that class by means of a QR decomposition
applied on the variables filtered by the Kalman filter.

\section{Bibliography}\label{bibliography-1}

ANSLEY F. C. and KOHN R. (1985), ``Estimating, filtering and smoothing
in state space models with incompletely specified initial conditions'',
The annals of statistics, vol.~13, n°4, 1286-1316.

DURBIN J. AND KOOPMAN S.J. (2012): ``Time Series Analysis by State Space
Methods'', second edition. Oxford University Press.

FRANCKE, M. K., KOOPMAN, S. J. and DE VOS, A. F. (2010), ``Likelihood
functions for state space models with diffuse initial conditions'',
Journal of Time Series Analysis, 31, 6.

KOHN R. AND ANSLEY F. C. (1985), ``Efficient estimation and prediction
in time series regression models'', Biometrika, 72, 694-697.

\chapter{Augmented filters}\label{augmented-filters}

The augmented filters are based on papers of De Jong et al.~See more
especially De Jong and Chu-Chun-Lin {[}2003{]}. So, when we talk about
the augmented Kalman filters (AKF), we talk about the solution proposed
by De Jong, Its implementation in JD+ contains a few modifications in
comparison with the reference paper.

\section{Filtering}\label{filtering}

A first approach consists in transforming the state space form into a
regression model. We build recursively the regression variables \(X\)
generated by the diffuse constraints and we apply the Kalman filter on
\(y, X\) to get \(y_L, X_L\). The least squares problem obtained that
way is then solved by a robust algorithm like the QR decomposition. See
the chapter on the computation of the likelihood for more details. The
solution is computed with all the data (no ``collapsing out'' of the
diffuse effects). Even if this solution is not the most parsimonious
one, it is usually quite robust.

A second approach consists in using the first data to get a rapid
estimate of the diffuse effects, in ``re-injecting'' that estimate in
the model and in finishing the processing with the ordinary Kalman
filter (the so-called augmented Kalman filter with a collapsing out of
the diffuse initialization). Usually, that collapsing is processed as
quickly as possible, though it is not mandatory. A better strategy would
consist in waiting till we have a robust enough estimate of the diffuse
effects (using for instance some condition number; see below for the
solution used in JD+).

So, in the first case, we consider

\[
\begin{align*}
B_t &= \prod{T_0...T_{t-1}}B \\
\text{or } B_t &= T_{t-1}B_{t-1} \quad, \text{with } B_0=B \\
X_t &= Z_t B_T
\end{align*}
\]

Applying the filter on \(y\) and on \(X\) gives

\[
\begin{align*}
e_t &= y_t-Z_t a_t \\
{y_L}_t &= e_t / \sqrt f_t \\
a_{t+1} &= T_t a_t + \tilde{K_t} {y_L}_t \\
E_t &= X_t - Z_t A_t = Z_t (B_t - A_t) \\ 
{X_L}_t &= (Z_t B_t - Z_t A_t) / \sqrt f_t \\
A_{t+1} &= T_t A_t + \tilde{K_t} {X_L}_t \quad \text{, with }A_0 = 0 
\end{align*}
\] Equivalently, we can use the following recursions

\[
\begin{align*}
\tilde{A_o} &= B_0 \\
\tilde{E_t} &= 0 -  Z_t \tilde A_t \\
\tilde {X_L}_t &= ( - Z_t \tilde A_t) / \sqrt f_t \\
\tilde A_{t+1} &= T_t \tilde A_t + \tilde{K_t} \tilde{X_L}_t
\end{align*}
\]

We can easily check by induction that:

\[
\begin{align*}
\tilde{A_t} &= B_t - A_t \\
\tilde E_t &= - E_t \\
\tilde {X_L}_t &= - {X_L}_t
\end{align*}
\]

De Jong used in his paper the second form, which is slightly faster. The
first form should be used if the marginal likelihood is needed (explicit
computation of \(X\))

The augmented filter is identical to the ordinary filter augmented by
the effects due to the diffuse initialization. Using the notations of
the ordinary filter, the new equations are:

\subsection{Update step t}\label{update-step-t-2}

In the univariate case, we add:

\[
\begin{align*}
E_t &= - Z_t A_t \\
A_{t|t} &= A_t + M_t f_t^{-1}E_t 
\end{align*}
\] In the multivariate case, we use:

\[
\begin{align*}
E_t &= - Z_t A_t \\
Ut &= R_t^{-1}E_t \\
A_{t|t} &= A_t + \tilde {M_t} U_t 
\end{align*}
\]

\subsection{Forecast step t}\label{forecast-step-t-2}

\[
\begin{align*}
A_{t+1} &= T_t A_{t|t} \\
\end{align*}
\]

\subsection{Q Matrix}\label{q-matrix}

The results of the previous iterations must be stored to compute the
usual statistics or additional results (likelihood, smoothing\ldots) In
the original paper, the so-called \(Q\) matrix ``accumulates'' the
filtered quantities and their interactions. It is defined recursively by

\[
Q_t = Q_{t-1} + (E_t, e_t)' {f_t}^{-1} (E_t, e_t) \quad , Q_{-1} = 0
\]

or

\[
Q_t = Q_{t-1} + (-\tilde{E_t}, e_t)' {f_t}^{-1} (-\tilde{E_t}, e_t) \quad , Q_{-1} = 0
\]

In a more compact form, it gives

\[
Q_t = Q_{t-1} + ( {X_L}_t, {y_l}_t)'( {X_L}_t, {y_l}_t)
\]

or

\[
Q_t = Q_{t-1} + (-\tilde {X_L}_t, {y_l}_t)'(-\tilde {X_L}_t, {y_l}_t)
\]

If we write

\[
Q_m = 
\begin{pmatrix}
S & s \\
s' & q
\end{pmatrix}
\]

we have that the ml estimator of \(\delta\) after \(m\) observations is
\(S^{-1}s\) and that the ml estimator of \(\sigma^2\) is \(q/m\) We
re-inject the effect of the diffuse initialization by adding to
\(a_m, P_m, q_m\) respectively
\(-A_m S^{-1}s, A_m S^{-1} A_m', s' S^{-1} s\)

To find the ``collapsing point'' \(m\), \(S\) must be invertible.
Several approaches can be considered to achieve that goal.

\begin{itemize}
\tightlist
\item
  Use directly the \(Q\) matrix and check the invertibility of \(S\) by
  Cholesky
\item
  Triangularize \(Q\) either completely or triangularize only \(S\) by
  means of (fast) Givens transformations, so that S in never explicitly
  computed
\item
  Accumulate \((\tilde {X_L}_t, {y_l}_t)\) in a matrix and use QR
  (recursively or not) to find the corresponding least squares problem
\item
  \ldots{}
\end{itemize}

See below some explanations on the triangularization of Q (t omitted).

We suppose that the triangular form of \(Q\) is \(Q_T\) so that
\(Q = Q_T Q_T'\)

\[
Q_T=\begin{pmatrix}
A & 0 \\
b & c
\end{pmatrix}
\]

We have that:

\[
\begin{align*}
S = AA' \\
s = A b' \\
q=b b' + c^2 \\
s' S^{-1} s = b A' (AA')^{-1} A b' = b b' \\
q - s' S^{-1} s = c^2 \\
S^{-1} s = (AA')^{-1} A b' = A'^{-1}b'
\end{align*}
\]

The ml estimator of \(\delta\) is then \(bA^{-1}\), its variance
\(\Psi= S^{-1}\) s and the ml estimator of \(\sigma^2\) is \(c^2/n\)

\subsection{Collapsing}\label{collapsing}

As mentioned above, the collapsing is performed as follow:

\[ 
\begin{align*}
a_{m_c} &= a_m - A_m \hat{\delta} \\
P_{m_c} &= P_m + A_m \Psi A_m'
\end{align*}
\]

\section{Smoothing}\label{smoothing}

\subsection{Computation of the
smoothations}\label{computation-of-the-smoothations}

\[ 
\begin{align*}
K_t &= T_t P_t Z_t' {f_t}^{-1} \\
\tilde e_t &=  f_t^{-1}e_t - K_t'r_t \\
\tilde E_t &=  f_t^{-1}E_t - K_t'R_t \\
var(\tilde e_t) &= f_t^{-1} + K_t' N_t K_t \\
var(\tilde E_t) &= f_t^{-1} + K_t' N_t K_t \\
\end{align*}
\]

\subsection{Main recursions}\label{main-recursions}

\[ 
\begin{align*}
r_{t-1}' &= \tilde e_t Z_t + r_t' T_t \\
R_{t-1}' &= \tilde E_t Z_t + R_t' T_t \\
L_t &= T_t - K_t Z_t \\
N_{t-1} &= Z_t' f_t^{-1} Z_t + L_t' N_t L_t 
\end{align*}
\]

\subsection{Update of the state}\label{update-of-the-state}

\[ 
\begin{align*}
&\text{using } B_t = A_t + P_t R_{t-1} \\
\hat \delta &= E(\delta \vert y) = S^{-1}s\\
\Psi &= var(\delta \vert y) = S^{-1} \\
\tilde a_t &= a_t + P_t r_{t-1} + B_t \hat \delta \\
\tilde P_t &= P_t - P_t N_{t-1} P_t +B_t \Psi B_t'
\end{align*}
\]

\subsection{\texorpdfstring{Update of the state in the case of
collapsing in
\(m\)}{Update of the state in the case of collapsing in m}}\label{update-of-the-state-in-the-case-of-collapsing-in-m}

\[ 
\begin{align*}
\hat \delta = E(\delta \vert y) &= S^{-1}(s + A'_m r_{m-1}) \\
\Psi = var(\delta \vert y) & = S^{-1} - S^{-1}(A_m' N_{m-1} A_m)S^{-1} \\
\end{align*}
\] \[ 
\begin{align*}
C_t &= P_t(R_{t-1} + N_{t-1} A_t) \\
\tilde a_t &= a_t + P_t r_{t-1} + B_t \hat \delta \\
\tilde P_t &= P_t - P_t N_{t-1} P_t + B_t \Psi B_t'- B_t S^{-1} C'_t - C_t S^{-1} B'_t
\end{align*}
\]

\chapter{DK filters}\label{dk-filters}

\part{Implementations}

\chapter{Java implementation}\label{java-implementation}

\chapter{R implementation}\label{r-implementation}

\section{Examples}\label{examples}

\subsection{Atomic blocks}\label{atomic-blocks-1}

\subsubsection{ARMA models}\label{arma-models}

\subsection{Models}\label{models}

\subsubsection{Basic structural model}\label{basic-structural-model}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{s}\OtherTok{\textless{}{-}}\NormalTok{rjd3toolkit}\SpecialCharTok{::}\NormalTok{ABS}\SpecialCharTok{$}\NormalTok{X0.}\DecValTok{2}\NormalTok{.}\DecValTok{20}\NormalTok{.}\FloatTok{10.}\NormalTok{M}
\NormalTok{model}\OtherTok{\textless{}{-}}\NormalTok{rjd3sts}\SpecialCharTok{::}\FunctionTok{model}\NormalTok{()}
\NormalTok{llt}\OtherTok{\textless{}{-}}\FunctionTok{locallineartrend}\NormalTok{(}\StringTok{"llt"}\NormalTok{)}
\NormalTok{seas}\OtherTok{\textless{}{-}}\FunctionTok{seasonal}\NormalTok{(}\StringTok{"seas"}\NormalTok{, }\AttributeTok{period=}\FunctionTok{frequency}\NormalTok{(s), }\AttributeTok{type=}\StringTok{"HarrisonStevens"}\NormalTok{)}
\NormalTok{noise}\OtherTok{\textless{}{-}}\FunctionTok{noise}\NormalTok{(}\StringTok{"n"}\NormalTok{, }\AttributeTok{fixed =} \ConstantTok{TRUE}\NormalTok{)}
\FunctionTok{add}\NormalTok{(model, llt)}
\FunctionTok{add}\NormalTok{(model, seas)}
\FunctionTok{add}\NormalTok{(model, noise)}
\NormalTok{rslt}\OtherTok{\textless{}{-}}\FunctionTok{estimate}\NormalTok{(model, }\FunctionTok{log}\NormalTok{(s))}
\end{Highlighting}
\end{Shaded}

\chapter{Rcpp implementation}\label{rcpp-implementation}




\end{document}
